---
title: "Hilbert Curve"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hilbert Curve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>"
)
```

In this article, we will create the Hilbert Curve:


```{r, message = F, echo = F}
library(animate)
device <- animate$new(600, 600, virtual = TRUE,
                      attr = list(style = "border:1px solid lightgray"))
attach(device)
set_max_stacksize(0)
clear()
#-------------------------------------------------------------------------

shift_x <- function(m0, k) { m0[,1] <- m0[,1] + k; return(m0); }
shift_y <- function(m0, k) { m0[,2] <- m0[,2] + k; return(m0); }
flip_45 <-  function(m0) {
    x_ref_origin <- m0[, 1] - min(m0[, 1])
    y_ref_origin <- m0[, 2] - min(m0[, 2])
    new_x <- min(m0[, 1]) + y_ref_origin
    new_y <- min(m0[, 2]) + x_ref_origin
    cbind(new_x, new_y)
}
flip_n45 <-  function(m0) {
    x_ref_origin <- m0[, 1] - min(m0[, 1])
    y_ref_origin <- m0[, 2] - min(m0[, 2])
    new_x <- min(m0[, 1]) - y_ref_origin + diff(range(m0[,2]))
    new_y <- min(m0[, 2]) - x_ref_origin + diff(range(m0[,1]))
    cbind(new_x, new_y)
}

hilbert <- function(n) {
    if (n == 1) {
        return(cbind(c(0.25, 0.25, 0.75, 0.75),
                     c(0.25, 0.75, 0.75, 0.25)))
    }
    if (n >= 2) {
        m1 <- hilbert(n-1) * 0.5
        k <- 0.5
        return(rbind(
            flip_45(m1),
            shift_y(m1, k),
            shift_x(shift_y(m1, k), k),
            flip_n45(shift_x(m1, k))
        ))
    }
}

mat <- hilbert(4)
par(xlim = range(mat[,1]), ylim = range(mat[,2]))
for (i in 1:nrow(mat)) {
    pts <- mat[i, ]
    points(pts[1], pts[2], cex = 50)
    if (i != 1) {
        last <- mat[i-1, ]
        lines(c(pts[1], last[1]),
              c(pts[2], last[2]))
    }
    # Sys.sleep(0.01)
}
rmd_animate(device, loop(1))
```




---

## 0. Setup

We will use the following function to aid the development. This function helps us set up the device and call the appropriate function to render and embed the animated plot into the R Markdown Document. (You may need to install the additional dependency of `pryr`.)

```{r, attr.source='.numberLines'}
animate_it <- function(..., options = click_to_loop()) {
  # Setup
  require(animate)   # 'require' is designed for use inside functions
  device <- animate$new(
    600, 600, virtual = TRUE,
    attr = list(style = "border:1px solid lightgray")
  )
  attach(device)
  clear()
  
  # Main code
  pryr::f(...)()
  
  # Embed animated plot in R Markdown Document
  rmd_animate(device, options)
}
```

#### Notes

- `virtual = TRUE` on line 5 and `rmd_animate(...)` on line 14 are only needed when the animation is developed in-line within the RMD file.






---

## 1. Hilbert curve, first order

The first order of Hilbert curve has only 4 points: (0.25,0.25), (0.25,0.75), (0.75, 0.75), (0.75, 0.25).

```{r}
hilbert <- function(n) {
    if (n == 1) {
        return(cbind(c(0.25, 0.25, 0.75, 0.75),
                     c(0.25, 0.75, 0.75, 0.25)))
    }
}
```

We will loop over the 4 points, draw them on the screen and connect each point to the previous point with a line.

```{r, message = F}
mat <- hilbert(1)

# Plotting the Hilbert curve
animate_it({
  # Set a fixed range
  par(xlim = c(0,1), ylim = c(0,1))
  
  # Loop over the data points
  for (i in 1:nrow(mat)) {
      pts <- mat[i, ]
      
      # Draw the point
      points(pts[1], pts[2], cex = 50)
      
      # Connect to the previous point with a line
      if (i != 1) {
          last <- mat[i-1, ]
          lines(c(pts[1], last[1]), c(pts[2], last[2]))
      }
  }
})  # Click on the canvas to run the animation
```

Note that the code for plotting will stay the same for the rest of this article! I would suggest to wrap that into a function that takes the `mat` variable as an argument. It should look something like:

```{r}
draw_hilbert <- function(mat) {
  animate_it({
    # copy the multiple lines of code from above
  })
}
```

```{r, echo = F}
draw_hilbert <- function(mat) {
  animate_it({
    par(xlim = range(mat[,1]), ylim = range(mat[,2]))
    for (i in 1:nrow(mat)) {
        pts <- mat[i, ]
        points(pts[1], pts[2], cex = 50)
        if (i != 1) {
            last <- mat[i-1, ]
            lines(c(pts[1], last[1]), c(pts[2], last[2]))
        }
    }
  })  
}
```




---

## 2. Hilbert curve, second order

### First step

As we go one order up, we will take the original Hilbert curve, copy it 4 times and lay it out at the 4 corners. In terms of (Cartesian) coordinates, 

- we scale the original coordinates by half to get the bottom-left corner piece;
- we shift the bottom-left corner piece to the top by half of the screen size to get the top-left corner;
- we shift the bottom-left corner piece to the right by half of the screen size to get the bottom-right corner;
- we shift the bottom-left corner piece to the top and right by half of the screen size to get the top-right corner;


Let's define two helper functions.

```{r}
shift_x <- function(m0, k) { m0[,1] <- m0[,1] + k; return(m0); }
shift_y <- function(m0, k) { m0[,2] <- m0[,2] + k; return(m0); }
```


Then we modify the Hilbert curve implementation.

```{r, attr.source='.numberLines'}
hilbert <- function(n) {
    if (n == 1) {
        return(cbind(c(0.25, 0.25, 0.75, 0.75),
                     c(0.25, 0.75, 0.75, 0.25)))
    }
    if (n >= 2) {
        k <- 0.5
        bottom_left <- hilbert(n-1) * 0.5
        top_left <- shift_y(bottom_left, k)
        top_right <- shift_x(top_left, k)
        bottom_right <- shift_x(bottom_left, k)
        return(rbind(bottom_left, top_left, top_right, bottom_right))
    }
}

mat <- hilbert(2)
draw_hilbert(mat)
```





### Second step

We notice that there are some crossover of lines, and they need to be fixed. There are several ways to do it. We will simply flip the bottom-left piece along the 45 degree line and the bottom-right piece by the -45 degree line.


To do this, we need two more helper functions.

```{r}
flip_45 <-  function(m0) {
    x_ref_origin <- m0[, 1] - min(m0[, 1])
    y_ref_origin <- m0[, 2] - min(m0[, 2])
    new_x <- min(m0[, 1]) + y_ref_origin
    new_y <- min(m0[, 2]) + x_ref_origin
    cbind(new_x, new_y)
}

flip_n45 <-  function(m0) {
    x_ref_origin <- m0[, 1] - min(m0[, 1])
    y_ref_origin <- m0[, 2] - min(m0[, 2])
    new_x <- min(m0[, 1]) - y_ref_origin + diff(range(m0[,2]))
    new_y <- min(m0[, 2]) - x_ref_origin + diff(range(m0[,1]))
    cbind(new_x, new_y)
}
```

And we modify line 12 of the Hilbert curve implementation.

```{r, attr.source='.numberLines'}
hilbert <- function(n) {
    if (n == 1) {
        return(cbind(c(0.25, 0.25, 0.75, 0.75),
                     c(0.25, 0.75, 0.75, 0.25)))
    }
    if (n >= 2) {
        k <- 0.5
        bottom_left <- hilbert(n-1) * 0.5
        top_left <- shift_y(bottom_left, k)
        top_right <- shift_x(top_left, k)
        bottom_right <- shift_x(bottom_left, k)
        return(rbind(flip_45(bottom_left), top_left, top_right, flip_n45(bottom_right)))
    }
}

mat <- hilbert(2)
draw_hilbert(mat)
```






---

## 3. Hilbert curve, higher order

By defining the base case (n = 1) and the recursive relation that reduces a higher order case into the lower order ones, we actually have all the cases for the natural number by "induction".

Here are the curves for the third order and the fourth order!

```{r, echo = T}
mat <- hilbert(3)
draw_hilbert(mat)
```

```{r, echo = T}
mat <- hilbert(4)
draw_hilbert(mat)
```





---

## 4. [Advanced] Transiting from one order to another order

Taking advantage of the `transition` argument, which can transit a set of points to another set of points (of the same size), we can transit from one Hilbert curve to another of higher order. To do this, we repeat the points in the lower order curve to match the number of points in the higher order curve. The full example is as follows:

```{r}
rep_mat <- function(mat, n) {
  matrix(rep(mat, each = n), nrow = nrow(mat) * n)
}

animate_it({
  par(xlim = c(0,1), ylim = c(0,1))
  
  max_order <- 6
  pts <- rep_mat(hilbert(1), 4^(max_order - 1))
  lines(pts[,1], pts[,2], id = "line-1")
  
  for (i in 2:max_order) {
    pts <- rep_mat(hilbert(i), 4^(max_order - i))
    lines(pts[,1], pts[,2], id = "line-1", transition = list(duration = 2000))
  }
}, options = click_to_play(start = 4))
```

