---
title: "Rose in Mathematics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rose in Mathematics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this article, we will make a mathematical rose like this:

```{r, echo = F, message=F}
rose <- function(theta, a = 1, k = 3) {
    # `a` controls the size, and `k` controls the shape of the rose
    list(x = a * cos(k * theta) * cos(theta),
         y = a * cos(k * theta) * sin(theta))
}
p <- 2
q <- 7
k <- p / q
is_odd <- function(x) x %% 2 == 1
period <- ifelse(is_odd(p) && is_odd(q), pi * q, 2 * pi * q)
time <- seq(from = 0, to = period, length.out = 400)  # 400 time steps
pts <- rose(time, a = 1, k = k)

library(animate)
border_with_round_corners <- "border:1px solid lightgray; border-radius:5px;"
device <- animate$new(600, 600, virtual = TRUE, attr = list(style = border_with_round_corners))
attach(device)
set_max_stacksize(0)
par(xlim = extendrange(pts$x), ylim = extendrange(pts$y))
for (step in seq_along(time)) {
    lines(pts$x[1:step], pts$y[1:step], id = "line-1")
    points(pts$x[step], pts$y[step], id = "point-1")
}
rmd_animate(device, loop(1))
```




---

## 1. Implement the `rose` function

The mathematical function that draws the rose is: 
$$(x,y) = \left(a \cos(kt) \cos(t), a \cos(kt) \sin(t)\right)$$
where $a$ and $k$ are the parameters controlling the size and the shape of the rose respectively.

```{r, exercise=TRUE}
rose <- function(theta, a = 1, k = 3) {
    # `a` controls the size, and `k` controls the shape of the rose
    list(x = a * cos(k * theta) * cos(theta), y = a * cos(k * theta) * sin(theta))
}
```

The parameter $k$ is specified with $k = p/q$, where $p$ and $q$ should be coprime, e.g. 1/3, 2/7, not 2/6, 3/6. If $p$ and $q$ are not coprime, then the rose will be drawn more than once.
```{r}
p <- 2
q <- 7
k <- p / q
```


Some calculation is needed to work out the number of time steps required to get to the end of the drawing of the rose. Assuming $p$ and $q$ are coprime, then the period is $\pi q$ if both $p$ and $q$ are odd, and $2 \pi q$ otherwise.

```{r, exercise=TRUE}
is_odd <- function(x) x %% 2 == 1
period <- ifelse(is_odd(p) && is_odd(q), pi * q, 2 * pi * q)
```


Now we compute the points on the rose with the parameters `a` and `k` for time=0 to time=`period`.

```{r}
time <- seq(from = 0, to = period, length.out = 400)  # 400 time steps
pts <- rose(time, a = 1, k = k)
```




---

## 2. Plot the Rose

### Set up the device

```{r, exercise=TRUE, message=FALSE, warning=FALSE, attr.source='.numberLines'}
library(animate)
border_with_round_corners <- "border:1px solid lightgray; border-radius:5px;"
device <- animate$new(600, 600, virtual = TRUE, attr = list(style = border_with_round_corners))
attach(device)
```

#### Notes

- line 1: `library(animate)` loads the `animate` library.

- line 2: We define a style to add a border around the plot so that we know it is there even when nothing has been drawn. The style is specified using CSS; if you haven't used it before, don't worry too much about that, the package website maintains a list of the most used ones for you to copy and paste to your code.

- line 3: `animate$new(600, 600, ...)` sets the canvas size to be 600 by 600 (width by height). 

- line 3: An additional flag `virtual = TRUE` is needed only for in-line RMD usage, i.e. you are animating the plot directly in the code chunk rather than importing it from a file.

- line 4: `attach(device)` lets us use the functions in `device` directly rather than typing `device$some_function` every time. To detach from the device, use `detach(device)` afterwards. 



### Making the plot

```{r, eval = F, attr.source='.numberLines'}
set_max_stacksize(0)
par(xlim = extendrange(pts$x), ylim = extendrange(pts$y))
for (step in seq_along(time)) {
    lines(pts$x[1:step], pts$y[1:step], id = "line-1")
    points(pts$x[step], pts$y[step], id = "point-1")
}
rmd_animate(device, click_to_loop(start = 3, wait = 10))
```

#### Notes 

- line 1: `set_max_stacksize(0)` sets the internal memory of the device to 0, so that it does not use any memory. 
- line 2: `par(xlim = ..., ylim = ...)` sets the ranges of the x-axis and the y-axis to be fixed at the specified values.
- line 3-6: `lines(...)` and `points(...)` draw lines and points on the screen. The argument `id` needs to be provided so that the graphics elements are updated, rather than redrawn, on the screen.
- line 7: this line is needed only for in-line RMD usage. It executes the animation using the virtual device and embeds the output in the RMD file.


```{r, eval = T, echo = F}
set_max_stacksize(0)
par(xlim = extendrange(pts$x), ylim = extendrange(pts$y))
for (step in seq_along(time)) {
    lines(pts$x[1:step], pts$y[1:step], id = "line-1")
    points(pts$x[step], pts$y[step], id = "point-1")
}
rmd_animate(device, click_to_loop(start = 3, wait = 10))
```


### Clean up 

If you are making the plot in the R console (not in RMD), you may want to `clear` the plot, switch `off` and detach the device when you are done.

```{r, eval=FALSE}
clear()
off()
detach(device)
```




---

## 3. [Advanced] Coordinating multiple plots

Continuing from above, we will draw a 4 x 4 grid of roses. First, we generate the roses using the previous code but looping through a range of $p$ and $q$ values.

```{r}
ps <- 1:4
qs <- c(5, 7, 11, 13)
roses <- list()
for (p in ps) {
  for (q in qs) {
    period <- ifelse(is_odd(p) && is_odd(q), pi * q, 2 * pi * q)
    time <- seq(from = 0, to = period, length.out = 400)
    roses[[length(roses) + 1]] <- rose(time, a = 1, k = p / q)
  }
}
```


Next, we create 15 additional plot areas using the `svg` function. Then when we want to make a plot at a specific area, we simply call `set(PLOT_AREA_ID)` just before the drawing commands. As the generated plot is quite large in size, we will do this directly in the R console.

```{r, message = F, eval = F, echo = T}
library(animate)
device <- animate$new(150, 150)
attach(device)
set_max_stacksize(0)


# Create 15 additional plot areas. The id of the first plot area is by default "SVG_1";
# we follow the convention and name the rest as "SVG_i", where i = 2, ..., 16.
for (i in 2:16) {
  svg(150, 150, id = paste0("SVG_", i))
}


# Loop over the plot area and draw the corresponding rose.
for (step in 1:400) {
  for (i in 1:16) {
    set(paste0("SVG_", i))  # Make plot area named "SVG_i" active
    pts <- roses[[i]]
    par(xlim = extendrange(pts$x), ylim = extendrange(pts$y))
    lines(pts$x[1:step], pts$y[1:step], id = "line-1")
    points(pts$x[step], pts$y[step], id = "point-1")
  }
}
```


<div>
<iframe width="600" height="600" id = "yt_embedded" src="https://www.youtube.com/embed/0hnEF11depo" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</div>

<script>document.querySelector("#yt_embedded").src = "https://www.youtube.com/embed/0hnEF11depo";</script>
