---
title: "Maze generation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maze generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Let's make a maze!

```{r, echo = F, message = F}
animate_it <- function(..., options = click_to_loop()) {
  # Setup
  require(animate)   # 'require' is designed for use inside functions
  device <- animate$new(680, 340, virtual = TRUE, 
                        attr = list(style = "border:1px solid lightgray"))
  attach(device)
  clear()
  
  # Main code
  pryr::f(...)()
  
  # Embed animated plot in R Markdown Document
  rmd_animate(device, options)
}


cell <- function(id, i, j, walls = rep(1, 4), visited = FALSE) {
  list(id = id, x = i, y = j, walls = walls, visited = visited)
}

cell_color <- \(x) ifelse(x, "green", "white")

draw_cells <- function(x, y, id, color) {
  bars(x = x, y = y, w = 1, h = 1, id = id, bg = color)
}


wall_color <- \(x) ifelse(x, "black", "green")
`%+%` <- paste0

draw_walls <- function(x, y, id, walls) {
    cw1 <- sapply(walls[[1]], wall_color)
    cw2 <- sapply(walls[[2]], wall_color)
    cw3 <- sapply(walls[[3]], wall_color)
    cw4 <- sapply(walls[[4]], wall_color)
    bars(  x, y+1, 1.1, 0.1, bg = cw1, id = "top-" %+% id) 
    bars(x+1,   y, 0.1, 1.1, bg = cw2, id = "right-" %+% id) 
    bars(  x,   y, 1.1, 0.1, bg = cw3, id = "bottom-" %+% id) 
    bars(  x,   y, 0.1, 1.1, bg = cw4, id = "left-" %+% id) 
}


visualise_maze <- function(cells, active_cell_id) {
  # Collect the data from the cells for plotting
  cx  <- sapply(cells, \(cell) cell$x)
  cy  <- sapply(cells, \(cell) cell$y)
  cid <- sapply(cells, \(cell) cell$id)
  cbg <- sapply(cells, \(cell) cell_color(cell$visited))
  cwalls <- lapply(1:4, \(wid) sapply(cells, \(cell) cell$walls[wid]))
                   
  cbg[which(cid == active_cell_id)] <- "blue"  # modified
  
  # Plot the cells
  draw_cells(cx, cy, cid, cbg)
  draw_walls(cx, cy, cid, cwalls)
}

Stack <- function() {
  env <- new.env()
  env$storage <- list()
  
  push <- function(x) {
    len <- length(env$storage)
    env$storage[[len + 1]] <- x  # add an item at the end
  }
  
  pop <- function() {
    len <- length(env$storage)
    last_item <- env$storage[[len]]   # retrieve the last item
    env$storage <- env$storage[-len]  # remove it from the sequence
    last_item                         # return the last item
  }
  
  peek <- function() {
    len <- length(env$storage)
    if (len == 0) NULL else env$storage[[len]]
  }
  
  is_empty <- function() {
    length(env$storage) == 0
  }
  
  list(push = push, pop = pop, peek = peek, is_empty = is_empty)
}


get_neighbors_id <- function(id, n, m) {
    # Set the value to be NA if the (x,y) coordinates go beyond the grid size
    clip <- function(coord) {
        x <- coord[1]
        y <- coord[2]
        c(ifelse(x > 0 && x <= m, x, NA),
          ifelse(y > 0 && y <= n, y, NA))
    }
    
    # Given an id, retrieve the (x,y) coordinates
    id_to_xy <- function(id) {
        c(x = 1 + (id - 1) %% m, y = 1 + (id - 1) %/% m)
    }
    
    # Convert the (x,y) coordinates to an id
    # - Return NA if any of the coordinates contains NA, indicating that the cell is outside of the grid.
    xy_to_id <- function(coord) {
        if (any(is.na(coord))) NA else (coord[2] - 1) * m + coord[1]
    }
    
    xy <- id_to_xy(id)
    c(top    = xy_to_id(clip(xy + c(0, 1))),
      right  = xy_to_id(clip(xy + c(1, 0))),
      bottom = xy_to_id(clip(xy + c(0, -1))),
      left   = xy_to_id(clip(xy + c(-1, 0))))
}


generate_maze <- function(n = 5, m = 10) {
    mat0 <- expand.grid(1:m, 1:n)
    cells <- lapply(1:nrow(mat0), \(i) cell(i, mat0[i, 1], mat0[i, 2]))
    cells[[1]]$visited <- TRUE

    # New implementation
    stack <- Stack()
    stack$push(cells[[1]])
    while (!stack$is_empty()) {
        # Plot the current cell
        current_cell <- stack$pop()
        visualise_maze(cells, current_cell$id)
        # Sys.sleep(0.1)  # needed if developing in the R console

        # Given the current cell, find the neighbours that have not been visited
        cid <- current_cell$id
        ids <- get_neighbors_id(cid, n, m)
        not_visited <- ids |>
            sapply(\(ind) if (is.na(ind)) NA else !cells[[ind]]$visited) |>
            which()
        
        # If there is at least one unvisited neighbour
        if (length(not_visited) >= 1) {
            # Push the current cell to the stack
            stack$push(current_cell)
          
            # Choose one of the unvisited neighbours
            pos <- ifelse(length(not_visited) == 1,
                          not_visited,
                          sample(not_visited, 1))
            chosen_id <- ids[pos]

            # Remove the walls between the current cell and the chosen cell.
            # Note that there are two walls to be removed, one is to exit
            # the current cell, the other is to enter the chosen cell.
            cells[[cid]]$walls[pos] <- 0
            
            rev_pos <- which(get_neighbors_id(chosen_id, n, m) == cid)
            cells[[chosen_id]]$walls[rev_pos] <- 0

            # Mark the chosen cell as visited and push it to the stack
            cells[[chosen_id]]$visited <- TRUE
            stack$push(cells[[chosen_id]])
        }
    }
    invisible(cells)
}


animate_it({
    n <- 10
    m <- 20
    par(xlim = c(1, m+1), ylim = c(1, n+1))
    generate_maze(n, m)
}, options = loop(1))
```





---

## 0. Setup

We will use the following function to aid the development. This function helps us set up the device and call the appropriate function to render and embed the animated plot into the R Markdown Document. (You may need to install the additional dependency of `pryr`.)

```{r, attr.source='.numberLines'}
animate_it <- function(..., options = click_to_loop()) {
  # Setup
  require(animate)   # 'require' is designed for use inside functions
  device <- animate$new(680, 340, virtual = TRUE, 
                        attr = list(style = "border:1px solid lightgray"))
  attach(device)
  clear()
  
  # Main code
  pryr::f(...)()
  
  # Embed animated plot in R Markdown Document
  rmd_animate(device, options)
}
```

#### Notes

- `virtual = TRUE` on line 5 and `rmd_animate(...)` on line 14 are only needed when the animation is developed in-line within the RMD file.




---

## 1. Make a cell and draw it

We will first define a cell. A cell has an `id`, the (`x`,`y`) coordinates, 4 walls (present or not) and an indicator to show if this cell has been visited before.

We paint the cell in green if the cell has been visited before and in white otherwise.

```{r, message = F}
cell <- function(id, i, j, walls = rep(1, 4), visited = FALSE) {
  list(id = id, x = i, y = j, walls = walls, visited = visited)
}

cell_color <- \(x) ifelse(x, "green", "white")

draw_cells <- function(x, y, id, color) {
  bars(x = x, y = y, w = 1, h = 1, id = id, bg = color)
}

animate_it({
  par(xlim = c(0, 20), ylim = c(0, 10))
  cell_1 <- cell("cell-1", 1, 1, visited = TRUE)
  draw_cells(cell_1$x, cell_1$y, cell_1$id, cell_color(cell_1$visited))
})
```


#### Add walls

Now we add the walls around the cells. The wall is painted in black if it is present and in green otherwise. We will design the function `draw_walls` in such a way that it can handle multiple cells.

```{r, message = F}
# Continue from above 
wall_color <- \(x) ifelse(x, "black", "green")
`%+%` <- paste0

draw_walls <- function(x, y, id, walls) {
    cw1 <- sapply(walls[[1]], wall_color)
    cw2 <- sapply(walls[[2]], wall_color)
    cw3 <- sapply(walls[[3]], wall_color)
    cw4 <- sapply(walls[[4]], wall_color)
    bars(  x, y+1, 1.1, 0.1, bg = cw1, id = "top-" %+% id) 
    bars(x+1,   y, 0.1, 1.1, bg = cw2, id = "right-" %+% id) 
    bars(  x,   y, 1.1, 0.1, bg = cw3, id = "bottom-" %+% id) 
    bars(  x,   y, 0.1, 1.1, bg = cw4, id = "left-" %+% id) 
}

animate_it({
  par(xlim = c(0, 20), ylim = c(0, 10))
  cell_1 <- cell("cell-1", 1, 1, visited = FALSE)
  draw_cells(cell_1$x, cell_1$y, cell_1$id, cell_color(cell_1$visited))
  draw_walls(cell_1$x, cell_1$y, cell_1$id, cell_1$walls)
})
```

---



## 2. Make a grid of cells

To make a grid of cells, we first generate a set of (x,y) coordinates, then initialise some cells using them. After the cells are initialised, we collect the relevant data and plot them on the screen using the `draw_cells` and `draw_walls` defined previously.

Note that in the `visualise_maze` below, we will include an optional argument `active_cell_id` to indicate where the active cell is. The active cell will be painted in blue, and this will be useful in the later sections.

```{r, message = F}
generate_maze <- function(m, n) {
  mat0 <- expand.grid(1:m, 1:n)
  cells <- lapply(1:nrow(mat0), \(i) cell(i, mat0[i, 1], mat0[i, 2]))
  cells[[1]]$visited <- TRUE
  visualise_maze(cells)
}

visualise_maze <- function(cells, active_cell_id) {
  # Collect the data from the cells for plotting
  cx  <- sapply(cells, \(cell) cell$x)
  cy  <- sapply(cells, \(cell) cell$y)
  cid <- sapply(cells, \(cell) cell$id)
  cbg <- sapply(cells, \(cell) cell_color(cell$visited))
  cwalls <- lapply(1:4, \(wid) sapply(cells, \(cell) cell$walls[wid]))
                   
  if (!missing(active_cell_id)) {
    cbg[which(cid == active_cell_id)] <- "blue"  
  }
  
  # Plot the cells
  draw_cells(cx, cy, cid, cbg)
  draw_walls(cx, cy, cid, cwalls)
}

animate_it({
  par(xlim = c(1, 21), ylim = c(1, 11))
  generate_maze(20, 10)
})
```




---

## 3. Make a maze!

We will use the randomized depth-first search algorithm [here](https://en.wikipedia.org/wiki/Maze_generation_algorithm) to generate the maze. The idea is to walk the grid randomly, and when encountered a dead-end, backtrack to the last (multi-way) intersection for the alternative path and continue until every cell in the grid has been visited.

### a. Make a stack

We will need the `Stack` data structure. A stack is a sequence of storages that follows the first-in-first-out protocol. It has two major operations, `push` and `pop`. `push` adds an element to the sequence, and `pop` removes the last added element from the sequence. We also add a `peek` operation to inspect the last element of the sequence without removing it from the storage and a `is_empty` operation to check if the stack is empty.

```{r}
Stack <- function() {
  env <- new.env()
  env$storage <- list()
  
  push <- function(x) {
    len <- length(env$storage)
    env$storage[[len + 1]] <- x  # add an item at the end
  }
  
  pop <- function() {
    len <- length(env$storage)
    last_item <- env$storage[[len]]   # retrieve the last item
    env$storage <- env$storage[-len]  # remove it from the sequence
    last_item                         # return the last item
  }
  
  peek <- function() {
    len <- length(env$storage)
    if (len == 0) NULL else env$storage[[len]]
  }
  
  is_empty <- function() {
    length(env$storage) == 0
  }
  
  list(push = push, pop = pop, peek = peek, is_empty = is_empty)
}

# Example
my_stack <- Stack()
my_stack$peek()
my_stack$is_empty()

my_stack$push(15)
my_stack$push(32)
my_stack$peek()

my_stack$pop()
my_stack$peek()
my_stack$is_empty()
```


### b. Search through the maze

To search through the maze, we begin at a particular cell, then randomly choose a neighbour cell that has not been visited and move to that cell. To do that, we will need a function `get_neighbors_id` that takes the id of the current cell and returns a list of its neighbour cells. (From there, we can check if each of the neighbours have been visited.)

Given the id of the cell, we retrieve its (x,y) coordinates, then simply add (-1,0), (1, 0), (0, -1), (0, 1) to get the four neighbour cells. Clipping may be needed when the cell is at the boundary of the grid, outside of which there is no cell. 

```{r}
get_neighbors_id <- function(id, n, m) {
    # Set the value to be NA if the (x,y) coordinates go beyond the grid size
    clip <- function(coord) {
        x <- coord[1]
        y <- coord[2]
        c(ifelse(x > 0 && x <= m, x, NA),
          ifelse(y > 0 && y <= n, y, NA))
    }
    
    # Given an id, retrieve the (x,y) coordinates
    id_to_xy <- function(id) {
        c(x = 1 + (id - 1) %% m, y = 1 + (id - 1) %/% m)
    }
    
    # Convert the (x,y) coordinates to an id
    # - Return NA if any of the coordinates contains NA, indicating that the cell is outside of the grid.
    xy_to_id <- function(coord) {
        if (any(is.na(coord))) NA else (coord[2] - 1) * m + coord[1]
    }
    
    xy <- id_to_xy(id)
    c(top    = xy_to_id(clip(xy + c(0, 1))),
      right  = xy_to_id(clip(xy + c(1, 0))),
      bottom = xy_to_id(clip(xy + c(0, -1))),
      left   = xy_to_id(clip(xy + c(-1, 0))))
}
```


Now, we implement the search algorithm in the `generate_maze` function.

```{r}
generate_maze <- function(n = 5, m = 10) {
    mat0 <- expand.grid(1:m, 1:n)
    cells <- lapply(1:nrow(mat0), \(i) cell(i, mat0[i, 1], mat0[i, 2]))
    cells[[1]]$visited <- TRUE

    # New implementation
    stack <- Stack()
    stack$push(cells[[1]])
    while (!stack$is_empty()) {
        # Plot the current cell
        current_cell <- stack$pop()
        visualise_maze(cells, active_cell_id = current_cell$id)
        # Sys.sleep(0.1)  # needed if developing in the R console

        # Given the current cell, find the neighbours that have not been visited
        cid <- current_cell$id
        ids <- get_neighbors_id(cid, n, m)
        not_visited <- ids |>
            sapply(\(ind) if (is.na(ind)) NA else !cells[[ind]]$visited) |>
            which()
        
        # If there is at least one unvisited neighbour
        if (length(not_visited) >= 1) {
            # Push the current cell to the stack
            stack$push(current_cell)
          
            # Choose one of the unvisited neighbours
            pos <- ifelse(length(not_visited) == 1,
                          not_visited,
                          sample(not_visited, 1))
            chosen_id <- ids[pos]

            # Remove the walls between the current cell and the chosen cell.
            # Note that there are two walls to be removed, one is to exit
            # the current cell, the other is to enter the chosen cell.
            cells[[cid]]$walls[pos] <- 0
            
            rev_pos <- which(get_neighbors_id(chosen_id, n, m) == cid)
            cells[[chosen_id]]$walls[rev_pos] <- 0

            # Mark the chosen cell as visited and push it to the stack
            cells[[chosen_id]]$visited <- TRUE
            stack$push(cells[[chosen_id]])
        }
    }
    invisible(cells)
}
```


Finally, let's run the maze!

```{r, message = F}
animate_it({
    n <- 10
    m <- 20
    par(xlim = c(1, m+1), ylim = c(1, n+1))
    generate_maze(n, m)
}, options = click_to_loop(wait = 10))
```
